Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-15 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Comments on current draft

\newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
%\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
\newcommand{\islock}{\boxdotright}
\newcommand{\lockvar}{\islock^{v}}
\newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
\newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\emp}{\mathsf{emp}}

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

@@

Iris~\cite{iris} is a framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''. Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state. Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity. Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}. All proofs described in this paper have been formalized in this VST+Iris setting.
%What else do we need to say?

@@

Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\:p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\:p\ (t[k \mapsto v])\}

@@

\{\mathsf{BST}\:p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\:p\ (t[k \mapsto \_])\}
\end{mathpar}
where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$. For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.

@@

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle

@@

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it. Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}. The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock. By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
 lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
$$\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}$$
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

@@

At the leaves, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

\subsubsection{Per-Node and Global Ghost State}
Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal. Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it. This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations. More precisely, the abstract state of a node in the tree is of type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$. The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ and contents $n$, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.

The $\treerep$ predicate describes the conditions under which an abstract map $m$ is implemented by the ghost state of a tree. Intuitively, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$. We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree:
$$\begin{array}{r l}
\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(l, r), \mathsf{None}}\\
& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(l, r), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
\end{array}$$
We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion. We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.

This collection of per-node ghost state describes the contents and structure of the tree, but a thread that acquires a node's lock also needs to know that the node it contains actually appears in the tree $t$. For this purpose, we add a ghost-state set $\mathsf{ghost\_nodes}(S)$ that records the identifiers of all nodes in the tree, defined so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$. Then our final definition of $\treerep$ is:
$$\begin{array}{r l}\treerep_g\ m, \triangleq &\exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty))\ * \\&\mathsf{ghost\_nodes}(\mathsf{ids}(t))\end{array}$$
The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$. Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node IDs and key-value pairs. We will take advantage of this property when verifying the \lstinline{delete} function.

>>>>

\newcommand{\LB}[1]{\textbf{\textcolor{red}{[ LB: #1]}}}
\newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
%\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
\newcommand{\islock}{\boxdotright}
\newcommand{\lockvar}{\islock^{v}}
\newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
\newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\emp}{\mathsf{emp}}

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that $P_l$ and $Q_l$ must be true before and after the call\LB{'Must be true': is there a partial-versus-total distinction here, or a reading where the precondition is an assumption (or implies safety) and the postoncdition is a guarantee?}, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. \LB{This suggests I should interpret the leftmost occurrence of 'a' as a binder whose scope extends all the way to $Q_p(a)$. So maybe the binding should happen outside the left pair of angle brackets? Also, as decorations 'l' and 'p' are not indices but just markers (they don't range over anything) and both 'private' and 'public' start with the letter 'p', maybe replacing $P_l$, $P_p$, $Q_l$, $Q_p$ with say $P$, $Q$, $R$, $S$, or maybe $P$, $P'$, $Q,$, $Q'$ could help. Indeed, the concrete patters below suggest something like $P$, $Q$, $P'$, $Q'$, to emphasize that the two predicates to the left of $\mid$ have something in common, and so do the two predicates to the right of $\mid$.} The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

@@

Iris~\cite{iris} is a framework that synthesizes ideas from a range of CSLs with the idea that ``ghost state is all you need''. Iris begins with a small core logic including arbitrary higher-order ghost state, and derives features including invariants and logical atomicity from particular instances of ghost state. Iris is designed as a language- and logic-independent framework, and recent versions of VST take advantage of this, incorporating Iris-style ghost state in the foundational model and directly using Iris tactics to reason about invariants and atomicity\LB{of actual C implementations}. Iris has also been used as the basis for proofs with atomic lock specifications, such as the work of Krishna et al.~\cite{templates}. All proofs described in this paper have been formalized in this VST+Iris setting.
%What else do we need to say?
\LB{Maybe swap the last 2 sentences?}

@@

\LB{Suppose versus: we have in VST's library, or ``will introduce below''}Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\:p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\:p\ (t[k \mapsto v])\}

@@

\{\mathsf{BST}\:p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\:p\ (t[k \mapsto \_])\}
\end{mathpar}
where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$. For instance, $t$ may be a recursively defined tree type in Coq\LB{It seems that $t$ is not a Coq type but a value of such a recursivle defined type. Maybe give the inductive type here.}, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.

@@

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). \LB{Here one wonders how the two predicates bst\_ref and bst\_abs are coordinated/related to each other, since their arguments are disjoint. Maybe just say that this wil be explained below, or is the magic of ghost state/names/identifiers?}Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.\LB{How does one create the first, empty tree? Maybe add a fourth operation/spec above?}

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure\LB{add comma here?} $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

We demonstrate our {\LB It's still not clear which one is `` ours'' and why it is prefereed\ldots} approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations\LB{As these specs were already given on page 5 we don't need to repeat (maybe put them in a Figure on page 5 and then refer to the figure here?}:
\begin{mathpar}
\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle

@@

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure\LB{Really? I had interpreted them on page 5 as being the C functions, as they were introduced in the sequential Hoare triples for BST}. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it. Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}. The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write\LB{label this definition as equation (1)} $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock. By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$ \LB{$\pi$ a rational(real?) number in the interval\ldots}, where the full share $\pi = 1$ is required to free the lock, while any \LB{nonzero?} share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
 lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
\LB{At first, I thought that the $\pi$ in $ \ell'' \islock_\pi R(n'')$ should/could be different from the one in $\ell \islock_\pi R(n)$. But the point is that its the same $\pi$, and $\pi$ is a parameter to entire setup. Maybe that's worth saying?}  the In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$ \LB{perhaps better:  \ldots we had to givegive up our knowledge of $n'$'s lock $\ell'$ in the second proof step (the call to release) here.}. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion. \LB{Do you mean some thing like Gotsman uses Locked(l), defined by $Locked(l) = l \mapsto_{\pi_1}R(n)$?}}\LB{i.e by replacing definition (1) by the following (and label the definition, too). And again $pi_1$ and $\pi_2$ are parameters, not chosen new each time.)}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent\LB{wait, locks have parents, too? I though just nodes in the tree}. Then the same sequence of operations gives us:
$$\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}$$
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible\LB{what does this mean, i.e. when/why might it not be possible?}, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

To verify our tree implementation\LB{I think here would be a good place to show the C structs and function prototypes of insert etc. (but not the function bodies)}, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. As described in section \ref{atomicity}, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

@@

At the leaves\LB{What's the difference between an empty node (eg the two leaves at the bottom of the figure) and a non-existent (leaf) node, ie one that could be the right child of node 40, for range (40,45)? Also, one of the two bounds of a range should be exclussive while the other one should be inclusive. Otherwise, eg the ranges of the two empty leaf nodes overlap as they both contain 35. (More generally, the ranges of any two children of a an inner node at oresent both contain the node's value, which seems to suggest that value could potentiall be found in three places).}, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

\subsubsection{Per-Node and Global Ghost State}
Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal. Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it. This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations. More precisely, the abstract state of a node in the tree is of \LB{Coq}type $\mathsf{range} \times \mathsf{option\ (Z \times val \times gname \times gname)}$, where the second component is $\mathsf{None}$ for a leaf node and $\mathsf{Some}\ (k, v, l, r)$ for an internal node with key $k$, value $v$, and left and right children $l$ and $r$. \LB{I think it would help to include the Coq definition of tree here and then decorate $t$, $g$, $l$, $r$ ith their Coq type (also, we use Z=key; not sure what the type of values is). And $l$ and $r$ are not children (which I'd interpret as trees) but gnames, a concept that hasn't been introduced yet}.The predicate $\nodeghost{g}{r, n}$ asserts that the node with identifier $g$ has range $r$ \LB{Now $r$ ranges over $Z \times Z)$ whereas the preceding sentence used $r$ to refer to a gname/child. And we should likely explain why nodes need identifiers in the first place - (the concept of ghost names is a technique/ artifact of Iris but to what extent is it visible at the top level? One wonders whether there's a bijecton between C-nodes in memory and gnames or some other data structure that keeps track of such an association, and how intensional the resulting reasoning style gets (invariance under renaming gnames...) )} and contents $n$ \LB{is ``content'' the same as the value range of the model-level KV store/BST?}, and will be divided between the lock invariant of the node and the abstract state assertion $\treerep$.

The $\treerep$ predicate describes the conditions under which an abstract map $m$ is implemented by the ghost state of a tree. Intuitively, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$. We formalize this by first defining the collection of per-node ghost state fragments needed to form an abstract tree\LB{node\_ghost has not yet been inyroduced. Is it a SL formula (ie mpred)? And now it seems $r$ refers to the upper bound of a range pair\ldots}:
$$\begin{array}{r l}
\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
& |\ \mathsf{Leaf} \Rightarrow \publichalf{g}{(l, r), \mathsf{None}}\\
& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \publichalf{g}{(l, r), \mathsf{Some}\ (k, v, g_l, g_r)}\ * \\&\qquad\mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
\end{array}$$
We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{node\_ghost}$ assertions of each node into a single assertion. We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.
